Drought indices of water demand and supply, soil moisture, vegetation, and its impact on LULCC in continental Chile

### 2024-01-19 article/paper_drought_Chile2023.qmd
actualizar articulo

# Study area (AGREGAR SUPERFICIE DE LANDCOVER EN CADA MACROZONA)

@@

@@

For the indices EDDI and SPEI that use water demand, first we have to calculate the AED. For this, we used the method of Hargreaves [@Hargreaves1994]:

@@

@@

To evaluate water demand, we chose the $EDDI$ [@Hobbins2016; @McEvoy2016] index, which uses the $AED$. For supply, we used the index recommended by the World Meteorological Organization (WMO) for monitoring drought, the SPI [@Mckee1993]. We calculated the SPEI, which used a balance between $P$ and $AED$, in this case, an auxiliary variable $D = P-AED$ is used. In this study, we used the $SSI$ (standardized soil moisture index at 1 m) [@Hao2013;@AghaKouchak2014], which uses SM. Finally, for the proxy of productivity, $zcNDVI$, we used the NDVI. Before using the NDVI, it was smoothed using a locally-weighted polynomial regression, following the procedure described in @Zambrano2018 and @Zambrano2016.

@@

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS. This product has a yearly frequency from 2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into ten macroclasses for forest, schrublands, savannas, grasslands, wetlands, croplands, urban, snow and ice, barren, and water bodies. To validate the IGBP MODIS, we compare the macroclasses with the ones of a more detailed landcover map made by @Zhao2016 for Chile for the years 2013–2014 (LCChile). The later has a spatial resolution of 30 m and three levels of defined classes; from those, we used level 1, which fits with the macroclasses derived from the IGBP MODIS. We chose the years 2013 and 2014 from MODIS IGBP because the LCChile was made with data acquired in 2013 and 2014. We resampled LCChile to the spatial resolution of the IGBP MODIS using the nearest neighbor method. Then, we took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCCHILE, IGP2013, and IGBP2014. Finally, we applied a confusion matrix and estimated the accuracy and F1 score as metrics of performance. The validation of IGBP2013 and IGBP2014 with LCChile reached similar metrics of performance, having an accuracy of 0.82 and a F1 score of 0.66 (see SS1).

Later, with the goal of analyzing vegetation areas less affected by landcover change, we calculated a raster mask for IGBP MODIS considering the classes that remain for more than 80% of the years (2001–2022), which allows us to identify the areas with no landcover change for the macroclasses. @fig-LCprop shows the summary of the proportion of surface per landcover class and macrozone, derived from the mask over continental Chile.

Further, we calculated the surface occupied per landcover class into the five macrozones ("Norte Grande" to "Austral") per year. After that, we calculated the trend's change in surface; we used the Sen' slope [@Sen1968] based on Mann-Kendall [@Kendall1975]. This way, we obtain a matrix of trends of 5 x 5 (macrozones x landcover).

We will explore if the trend in landcover classess is associated to trend of the drought indices. For this we will use the regularizations techniques of Lasso [@Tibshirani2010] and Ridge regressión [@Hoerl1970]. Also, we will test random forest for this purpose [@Ho1995].

@@

To estimate if there are significant positive or negative trends for the drought indices, we used the non-parametric test of Mann-Kendall [@Kendall1975]. To determine the magnitude of the trend, we used Sen's slope [@Sen1968]. One of the advantages of applying this methodology is that the Sen's slope is not affected by outliers as regular regression does. We applied both to the six time scales from 1981 to 2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and SSI. In the case of zcNDVI (six months) was for 2000 to 2023. Thus, we have 31 trends. Also, we extracted the trend aggregated by macrozone and landcover class, obtaining a table of 31x5x5 (drought indices trends x macrozone x landcover class). We will use this data in @sec-methods_lulc to analyze if there is a strong relationship between the trends of drought indices and land cover surface within continental Chile.

For zcNDVI, we want to remove the effect of the Andes's mountain and mask out trends that were at an elevation above 1500m because there is scarce vegetation.

@@

@fig-zcNDVI_var and Table \ref{tab-landcoverTrend} show that the macrozones with major LULCC for 2001-2022 were "Centro", "Sur", and "Austral" with 36%, 31%, and 34%, respectively. The "Norte Chico" shows an increase in barrend land of 111 $km^2 year^{-1}$ and a reduction in the class savanna -70 $km^2 year^{-1}$. In the "Centro" and "Sur" there are changes in the Chilean matorral, with an important reduction in savanna (-136 to -318 $km^2 yr^{-1}$), and an increase in shrubland and grassland. Showing a change for more dense vegetation types. It appears to be a change from the surface occupied by cropland from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$) in the "Sur," replacing the savanna lost.

@@

@@

## Validation of ERA5L variables

>>>>

# Study area

@@

POR COMPLETAR

@@

For the indices EDDI and SPEI that use water demand, first we have to calculate the AED. For this, we used the method of Hargreaves [@Hargreaves1994;@Hargreaves1985]:

@@

We chose the method of Hargreaves to estimate AED because of its simplicity, which only requires temperatures and extrarrestrial radiation. Also, it has been recommended over other methods when the use of several climatic variables is limited [@Vicente-Serrano2014].

@@

To evaluate water demand, we chose the $EDDI$ [@Hobbins2016; @McEvoy2016] index, which uses the $AED$. For supply, we used the index recommended by the World Meteorological Organization (WMO) for monitoring drought, the SPI [@Mckee1993]. We calculated the SPEI, which used a balance between $P$ and $AED$, in this case, an auxiliary variable $D = P-AED$ is used. In this study, we used the $SSI$ (standardized soil moisture index at 1 m) [@Hao2013;@AghaKouchak2014], which uses soil moisture at 1m depth. Finally, for the proxy of productivity, $zcNDVI$, we used the NDVI. Before using the NDVI, it was smoothed using a locally-weighted polynomial regression, following the procedure described in @Zambrano2018 and @Zambrano2016.

@@

### Landcover macroclasess and validation

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS. This product has a yearly frequency from 2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies. Thus, we have a landcover raster time-series with the ten classes for 2001 and 2023.

To validate the landcover obtained, we compare the macroclasses with the ones of a more detailed landcover map made by @Zhao2016 for Chile with samples acquired in the years 2013–2014 (LCChile). The later has a spatial resolution of 30 m and three levels of defined classes; from those, we used level 1, which fits with the macroclasses landcover. We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from landcover macrolcasses to validate with LCChile.

We follow the next procedure:

i) resampled LCChile to the spatial resolution (500m) of the landcover macroclasses using the nearest neighbor method,
ii) took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCChile, IGBP2013, and IGBP2014; we considered the point extracted from LCChile as the truth and the values as the other two years as prediction
iii) calculate a confusion matrix with the classes extracted in the 1000 poitns for LCChile, IGBP2013, and IGBP2014. Calculate the performance metrics of accuracy and F1.

$$Accuracy = \frac{TP+TN}{TP+TN+FP+FN}=\frac{correct\,\, classifications}{all\,\, classifications}$$
$$F1=\frac{2\cdot TP}{2\cdot TP + FP +FN}$$
where $TP$ and $FN$ refer to true positive and false negative, correctly classified classes; $TN$ and $FP$ to true negative and false positive, wrongly classified classes.

### Landcover persistence mask 2001-2022

The time series of NDVI is affected by climatic conditions, vegetation development, seasonality, and changes in vegetation type. In this study, we want to analyze the variation in vegetation productivity in different landcover types and how it is affected by water demand, water supply, and soil moisture. In order to avoid changes due to a change in the landcover type, that will wrongly impact NDVI. We will develop a persistence mask for landcover for 2001–2023. Thereby, we reduce an important source of variation on a regional scale.

Thus, we calculated a raster mask for IGBP MODIS considering the macroclasses that remain without change for more than 80% of the years (2001–2022) per pixel, which allows us to identify the areas with no landcover change for the macroclasses.

### Landcover trend and drought indices

We calculated the surface occupied per landcover class into the five macrozones ("Norte Grande" to "Austral") per year for 2001–2023. After that, we calculated the trend's change in surface; we used the Sen' slope [@Sen1968] based on Mann-Kendall [@Kendall1975]. This way, we obtain a matrix of trends of 5 x 5 (macrozones x landcover). The aim is to later explore if the trend in landcover classes is associated with a trend in the drought indices. For this, we will use the techniques of regresion and regularization of Lasso [@Tibshirani2010] and Ridge [@Hoerl1970]. Also, we will test random forests for this purpose [@Ho1995]. We will choose the trend of landcover surface per macroclass and macrozone as the response variable and the trend of the drought indices (SPI, SPEI, EDDI, and SSI for time scales 1, 3, 6, 12, 24, and 36 months) as the predictor variables. With this analysis, we expect to gather insights regarding whether there is a pattern of climatic influence along Chile or if what is happening in Central Chile has to do with more localized climatic conditions.

@@

### Mann-Kendall and Sen's slope

To estimate if there are significant positive or negative trends for the drought indices, we used the non-parametric test of Mann-Kendall [@Kendall1975]. To determine the magnitude of the trend, we used Sen's slope [@Sen1968]. Some of the advantages of applying this methodology are that the Sen's slope is not affected by outliers as regular regression does, and it is a non-paramteric method that is not affected by the distribution of the data. We applied both to the six time scales from 1981 to 2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and SSI. In the case of zcNDVI (six months) was for 2000 to 2023. Thus, we have 31 trends. Also, we extracted the trend aggregated by macrozone and landcover class, obtaining a table of 31x5x5 (drought indices trends x macrozone x landcover class). We will use this data in @sec-methods_lulc to analyze if there is a strong relationship between the trends of drought indices and land cover surface within continental Chile.

### Trend in vegetation productivity without landcover change

@@

### Landcover macroclasess and validation

The validation of IGBP2013 and IGBP2014 with LCChile reached similar metrics of performance, having an accuracy of 0.82 and a F1 score of 0.66 (see SS1).

### Landcover persistence mask 2001-2022

The macrozones with major LULCC for 2001-2022 were "Centro", "Sur", and "Austral" with 36%, 31%, and 34%, respectively (@fig-studyArea and Table \ref{tab-landcoverTrend})

@fig-LCprop shows the summary of the proportion of surface per landcover class and macrozone, derived from the mask over continental Chile.

### Landcover trend and drought indices

@@

The "Norte Chico" shows an increase in barrend land of 111 $km^2 year^{-1}$ and a reduction in the class savanna -70 $km^2 year^{-1}$. In the "Centro" and "Sur" there are changes in the Chilean matorral, with an important reduction in savanna (-136 to -318 $km^2 yr^{-1}$), and an increase in shrubland and grassland. Showing a change for more dense vegetation types. It appears to be a change from the surface occupied by cropland from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$) in the "Sur," replacing the savanna lost.

@@

### 2024-01-19 article/paper_drought_Chile2023.tex
actualizar articulo

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

\hypertarget{study-area-agregar-superficie-de-landcover-en-cada-macrozona}{%
\section{Study area (AGREGAR SUPERFICIE DE LANDCOVER EN CADA
MACROZONA)}\label{study-area-agregar-superficie-de-landcover-en-cada-macrozona}}

@@

For the indices EDDI and SPEI that use water demand, first we have to
calculate the AED. For this, we used the method of Hargreaves
\citep{Hargreaves1994}:

@@

@@

To evaluate water demand, we chose the \(EDDI\)
\citep{Hobbins2016, McEvoy2016} index, which uses the \(AED\). For
supply, we used the index recommended by the World Meteorological
Organization (WMO) for monitoring drought, the SPI \citep{Mckee1993}. We
calculated the SPEI, which used a balance between \(P\) and \(AED\), in
this case, an auxiliary variable \(D = P-AED\) is used. In this study,
we used the \(SSI\) (standardized soil moisture index at 1 m)
\citep{Hao2013, AghaKouchak2014}, which uses SM. Finally, for the proxy
of productivity, \(zcNDVI\), we used the NDVI. Before using the NDVI, it
was smoothed using a locally-weighted polynomial regression, following
the procedure described in \citet{Zambrano2018} and
\citet{Zambrano2016}.

@@

@@

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
6.1 from MODIS. This product has a yearly frequency from 2001 to 2022.
The IGBP defines 17 classes; from these, we regrouped into ten
macroclasses for forest, schrublands, savannas, grasslands, wetlands,
croplands, urban, snow and ice, barren, and water bodies. To validate
the IGBP MODIS, we compare the macroclasses with the ones of a more
detailed landcover map made by \citet{Zhao2016} for Chile for the years
2013--2014 (LCChile). The later has a spatial resolution of 30 m and
three levels of defined classes; from those, we used level 1, which fits
with the macroclasses derived from the IGBP MODIS. We chose the years
2013 and 2014 from MODIS IGBP because the LCChile was made with data
acquired in 2013 and 2014. We resampled LCChile to the spatial
resolution of the IGBP MODIS using the nearest neighbor method. Then, we
took a random sample of 1000 points within continental Chile and
extracted the classes that fell within each point for LCCHILE, IGP2013,
and IGBP2014. Finally, we applied a confusion matrix and estimated the
accuracy and F1 score as metrics of performance. The validation of
IGBP2013 and IGBP2014 with LCChile reached similar metrics of
performance, having an accuracy of 0.82 and a F1 score of 0.66 (see
SS1).

Later, with the goal of analyzing vegetation areas less affected by
landcover change, we calculated a raster mask for IGBP MODIS considering
the classes that remain for more than 80\% of the years (2001--2022),
which allows us to identify the areas with no landcover change for the
macroclasses. Figure~\ref{fig-LCprop} shows the summary of the
proportion of surface per landcover class and macrozone, derived from
the mask over continental Chile.

Further, we calculated the surface occupied per landcover class into the
five macrozones (``Norte Grande'' to ``Austral'') per year. After that,
we calculated the trend's change in surface; we used the Sen' slope
\citep{Sen1968} based on Mann-Kendall \citep{Kendall1975}. This way, we
obtain a matrix of trends of 5 x 5 (macrozones x landcover).

We will explore if the trend in landcover classess is associated to
trend of the drought indices. For this we will use the regularizations
techniques of Lasso \citep{Tibshirani2010} and Ridge regressión
\citep{Hoerl1970}. Also, we will test random forest for this purpose
\citep{Ho1995}.

To estimate if there are significant positive or negative trends for the
drought indices, we used the non-parametric test of Mann-Kendall
\citep{Kendall1975}. To determine the magnitude of the trend, we used
Sen's slope \citep{Sen1968}. One of the advantages of applying this
methodology is that the Sen's slope is not affected by outliers as
regular regression does. We applied both to the six time scales from
1981 to 2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and
SSI. In the case of zcNDVI (six months) was for 2000 to 2023. Thus, we
have 31 trends. Also, we extracted the trend aggregated by macrozone and
landcover class, obtaining a table of 31x5x5 (drought indices trends x
macrozone x landcover class). We will use this data in
Section~\ref{sec-methods_lulc} to analyze if there is a strong
relationship between the trends of drought indices and land cover
surface within continental Chile.

For zcNDVI, we want to remove the effect of the Andes's mountain and
mask out trends that were at an elevation above 1500m because there is
scarce vegetation.

Figure~\ref{fig-zcNDVI_var} and Table \ref{tab-landcoverTrend} show that
the macrozones with major LULCC for 2001-2022 were ``Centro'', ``Sur'',
and ``Austral'' with 36\%, 31\%, and 34\%, respectively. The ``Norte
Chico'' shows an increase in barrend land of 111 \(km^2 year^{-1}\) and
a reduction in the class savanna -70 \(km^2 year^{-1}\). In the
``Centro'' and ``Sur'' there are changes in the Chilean matorral, with
an important reduction in savanna (-136 to -318 \(km^2 yr^{-1}\)), and
an increase in shrubland and grassland. Showing a change for more dense
vegetation types. It appears to be a change from the surface occupied by
cropland from the ``Centro'' to the ``Sur.'' Also, there is a high
increase in forest (397 \(km^2 yr^{-1}\)) in the ``Sur,'' replacing the
savanna lost.

@@

@@

\hypertarget{validation-of-era5l-variables-1}{%
\subsection{Validation of ERA5L
variables}\label{validation-of-era5l-variables-1}}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, frame hidden, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi

@@

\hypertarget{study-area}{%
\section{Study area}\label{study-area}}

@@

For the indices EDDI and SPEI that use water demand, first we have to
calculate the AED. For this, we used the method of Hargreaves
\citep{Hargreaves1994, Hargreaves1985}:

@@

We chose the method of Hargreaves to estimate AED because of its
simplicity, which only requires temperatures and extrarrestrial
radiation. Also, it has been recommended over other methods when the use
of several climatic variables is limited \citep{Vicente-Serrano2014}.

@@

To evaluate water demand, we chose the \(EDDI\)
\citep{Hobbins2016, McEvoy2016} index, which uses the \(AED\). For
supply, we used the index recommended by the World Meteorological
Organization (WMO) for monitoring drought, the SPI \citep{Mckee1993}. We
calculated the SPEI, which used a balance between \(P\) and \(AED\), in
this case, an auxiliary variable \(D = P-AED\) is used. In this study,
we used the \(SSI\) (standardized soil moisture index at 1 m)
\citep{Hao2013, AghaKouchak2014}, which uses soil moisture at 1m depth.
Finally, for the proxy of productivity, \(zcNDVI\), we used the NDVI.
Before using the NDVI, it was smoothed using a locally-weighted
polynomial regression, following the procedure described in
\citet{Zambrano2018} and \citet{Zambrano2016}.

@@

\hypertarget{landcover-macroclasess-and-validation}{%
\subsubsection{Landcover macroclasess and
validation}\label{landcover-macroclasess-and-validation}}

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
6.1 from MODIS. This product has a yearly frequency from 2001 to 2022.
The IGBP defines 17 classes; from these, we regrouped into ten
macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9
to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands,
13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
Thus, we have a landcover raster time-series with the ten classes for
2001 and 2023.

To validate the landcover obtained, we compare the macroclasses with the
ones of a more detailed landcover map made by \citet{Zhao2016} for Chile
with samples acquired in the years 2013--2014 (LCChile). The later has a
spatial resolution of 30 m and three levels of defined classes; from
those, we used level 1, which fits with the macroclasses landcover. We
chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from landcover
macrolcasses to validate with LCChile.

We follow the next procedure:

\begin{enumerate}
\def\labelenumi{\roman{enumi})}
\tightlist
\item
  resampled LCChile to the spatial resolution (500m) of the landcover
  macroclasses using the nearest neighbor method,
\item
  took a random sample of 1000 points within continental Chile and
  extracted the classes that fell within each point for LCChile,
  IGBP2013, and IGBP2014; we considered the point extracted from LCChile
  as the truth and the values as the other two years as prediction
\item
  calculate a confusion matrix with the classes extracted in the 1000
  poitns for LCChile, IGBP2013, and IGBP2014. Calculate the performance
  metrics of accuracy and F1.
\end{enumerate}

\[Accuracy = \frac{TP+TN}{TP+TN+FP+FN}=\frac{correct\,\, classifications}{all\,\, classifications}\]
\[F1=\frac{2\cdot TP}{2\cdot TP + FP +FN}\] where \(TP\) and \(FN\)
refer to true positive and false negative, correctly classified classes;
\(TN\) and \(FP\) to true negative and false positive, wrongly
classified classes.

\hypertarget{landcover-persistence-mask-2001-2022}{%
\subsubsection{Landcover persistence mask
2001-2022}\label{landcover-persistence-mask-2001-2022}}

The time series of NDVI is affected by climatic conditions, vegetation
development, seasonality, and changes in vegetation type. In this study,
we want to analyze the variation in vegetation productivity in different
landcover types and how it is affected by water demand, water supply,
and soil moisture. In order to avoid changes due to a change in the
landcover type, that will wrongly impact NDVI. We will develop a
persistence mask for landcover for 2001--2023. Thereby, we reduce an
important source of variation on a regional scale.

Thus, we calculated a raster mask for IGBP MODIS considering the
macroclasses that remain without change for more than 80\% of the years
(2001--2022) per pixel, which allows us to identify the areas with no
landcover change for the macroclasses.

\hypertarget{landcover-trend-and-drought-indices}{%
\subsubsection{Landcover trend and drought
indices}\label{landcover-trend-and-drought-indices}}

We calculated the surface occupied per landcover class into the five
macrozones (``Norte Grande'' to ``Austral'') per year for 2001--2023.
After that, we calculated the trend's change in surface; we used the
Sen' slope \citep{Sen1968} based on Mann-Kendall \citep{Kendall1975}.
This way, we obtain a matrix of trends of 5 x 5 (macrozones x
landcover). The aim is to later explore if the trend in landcover
classes is associated with a trend in the drought indices. For this, we
will use the techniques of regresion and regularization of Lasso
\citep{Tibshirani2010} and Ridge \citep{Hoerl1970}. Also, we will test
random forests for this purpose \citep{Ho1995}. We will choose the trend
of landcover surface per macroclass and macrozone as the response
variable and the trend of the drought indices (SPI, SPEI, EDDI, and SSI
for time scales 1, 3, 6, 12, 24, and 36 months) as the predictor
variables. With this analysis, we expect to gather insights regarding
whether there is a pattern of climatic influence along Chile or if what
is happening in Central Chile has to do with more localized climatic
conditions.

\hypertarget{mann-kendall-and-sens-slope}{%
\subsubsection{Mann-Kendall and Sen's
slope}\label{mann-kendall-and-sens-slope}}

To estimate if there are significant positive or negative trends for the
drought indices, we used the non-parametric test of Mann-Kendall
\citep{Kendall1975}. To determine the magnitude of the trend, we used
Sen's slope \citep{Sen1968}. Some of the advantages of applying this
methodology are that the Sen's slope is not affected by outliers as
regular regression does, and it is a non-paramteric method that is not
affected by the distribution of the data. We applied both to the six
time scales from 1981 to 2023 (monthly frequency) and the indices SPI,
EDDI, SPEI, and SSI. In the case of zcNDVI (six months) was for 2000 to
2023. Thus, we have 31 trends. Also, we extracted the trend aggregated
by macrozone and landcover class, obtaining a table of 31x5x5 (drought
indices trends x macrozone x landcover class). We will use this data in
Section~\ref{sec-methods_lulc} to analyze if there is a strong
relationship between the trends of drought indices and land cover
surface within continental Chile.

\hypertarget{trend-in-vegetation-productivity-without-landcover-change}{%
\subsubsection{Trend in vegetation productivity without landcover
change}\label{trend-in-vegetation-productivity-without-landcover-change}}

\hypertarget{data-1}{%
\subsection{Data}\label{data-1}}

\hypertarget{validation-of-era5l-variables-1}{%
\subsubsection{Validation of ERA5L
variables}\label{validation-of-era5l-variables-1}}

\hypertarget{landcover-macroclasess-and-validation-1}{%
\subsubsection{Landcover macroclasess and
validation}\label{landcover-macroclasess-and-validation-1}}

The validation of IGBP2013 and IGBP2014 with LCChile reached similar
metrics of performance, having an accuracy of 0.82 and a F1 score of
0.66 (see SS1).

\hypertarget{landcover-persistence-mask-2001-2022-1}{%
\subsubsection{Landcover persistence mask
2001-2022}\label{landcover-persistence-mask-2001-2022-1}}

The macrozones with major LULCC for 2001-2022 were ``Centro'', ``Sur'',
and ``Austral'' with 36\%, 31\%, and 34\%, respectively
(Figure~\ref{fig-studyArea} and Table \ref{tab-landcoverTrend})

@@

Figure~\ref{fig-LCprop} shows the summary of the proportion of surface
per landcover class and macrozone, derived from the mask over
continental Chile.

\hypertarget{landcover-trend-and-drought-indices-1}{%
\subsubsection{Landcover trend and drought
indices}\label{landcover-trend-and-drought-indices-1}}

@@

The ``Norte Chico'' shows an increase in barrend land of 111
\(km^2 year^{-1}\) and a reduction in the class savanna -70
\(km^2 year^{-1}\). In the ``Centro'' and ``Sur'' there are changes in
the Chilean matorral, with an important reduction in savanna (-136 to
-318 \(km^2 yr^{-1}\)), and an increase in shrubland and grassland.
Showing a change for more dense vegetation types. It appears to be a
change from the surface occupied by cropland from the ``Centro'' to the
``Sur.'' Also, there is a high increase in forest (397 \(km^2 yr^{-1}\))
in the ``Sur,'' replacing the savanna lost.

@@
